---
title: "Lewiston Police Budgeting Overview v1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Introduction

In an effort to better understand how Lewiston's police budget compares to those in the rest of the state, I visualized the police budgets of select Maine towns. My initial data set contained the 12 most populous Maine towns, but due to difficulty obtaining strong data for 3 towns (Sanford, Scarborough and Saco), the final list of 9 towns became:

* Portland
* Lewiston
* Bangor
* South Portland
* Auburn
* Biddeford
* Brunswick
* Westbrook
* Augusta

All budget data was obtained via the respective town's website, and all other data was found on census.gov.

For the sake of this project, the 'budget' is defined as a town's municipal general fund expenses (without the school budget).

## Police Spending as a Percentage of Budget

```{r echo= FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(ggplot2)
library(readxl)
library(ggthemes)
library(ggrepel)
library(directlabels)
options(scipen = 100)
police_budgeting <- read_excel("~/projects/police/police-budgeting-2.xlsx")
####### load in Excel file

data <- subset(police_budgeting, select = -c(...3))
####### erase empty column

data$`Approved General Fund Budget` <- as.numeric(data$`Approved General Fund Budget`)
data$`Police Budget` <- as.numeric(data$`Police Budget`)
data$`% on Police` <- as.numeric(data$`% on Police`)
data$`Population (est.)` <- as.numeric(data$`Population (est.)`)
data$`% Change 2017-2021` <- as.numeric(data$`% Change 2017-2021`)
data$`% Change 2012-2021` <- as.numeric(data$`% Change 2012-2021`)
data$`Median Income` <- as.numeric(data$`Median Income`)
data$`Police $ per income` <-as.numeric(data$`Police $ per income`)
data$`Police$ per Citizen` <- as.numeric(data$`Police$ per Citizen`)
####### convert character columns into numeric

Lewiston_data <- data %>% filter(City == 'Lewiston')
###### create Lewiston dataset

years <- c (2010:2022)
###### create vector of years

Lewiston_data %>%
  ggplot(aes(`Fiscal Year`, `% on Police`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years) +
  scale_y_continuous("% of Budget spent on Police", limits = c(0.05, 0.20)) +
  stat_smooth(method = "lm", geom = "line", alpha = 0.75, size = 1, se = FALSE) +
  geom_point() +
  ggtitle("Police Spending in Lewiston") +
  theme_solarized_2() +
  scale_colour_manual(values = c("#dc322f"))
###### work in progress - Lewiston Police Budget 
```

In this first plot, we see the percentages of Lewiston's budget allocated to the police department from fiscal year 2010 to 2022. In FY 2010, 12.4% of the budget was allocated to the police and in FY 2022, 14.7%. 

## Adding Other Towns for Context
### Portland
```{r echo= FALSE, message = FALSE, warning = FALSE}
Portland_Lewiston_Data <- data %>% filter(City == 'Lewiston' | City == 'Portland')
###### create Portland/Lewiston dataset

Portland_Lewiston_Data %>%
  ggplot(aes(`Fiscal Year`, `% on Police`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years) +
  scale_y_continuous("% of Budget spent on Police", limits = c(0.05, 0.20)) +
  stat_smooth(method = "lm", geom = "line", alpha = 0.75, size = 1, se = FALSE) +
  geom_point() +
  ggtitle("Police Spending in Lewiston/Portland") +
  theme_solarized_2() +
  scale_colour_manual(values = c("#dc322f", "#268bd2"))
##### work in progress - Lewiston/Portland Police Budget
```

In order to better contextualize these percentages, we add other towns to the plot.

First, here is Lewiston's spending as compared to Portland's. Portland allocated 8.5% of their budget to police in FY 2010 and hovered around that mark for the rest of the decade. Notice the rather sharp decline in FY 2022 spending - down to 7%.

We can shrink our scope to fiscal years 2017-2022 to better highlight the opposite trajectories of Maine's two most populous towns

```{r echo= FALSE, message = FALSE, warning = FALSE}
Portland_Lewiston_Data %>%
  filter('Fiscal Year' >= 2017) %>%
  ggplot(aes(`Fiscal Year`, `% on Police`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years, limits = c(2017,2022)) +
  scale_y_continuous("% of Budget spent on Police", limits = c(0.05, 0.20)) +
  stat_smooth(method = "lm", geom = "line", alpha = 0.75, size = 1, se = FALSE) +
  geom_point() +
  ggtitle("Police Spending in Lewiston/Portland, last 5 years") +
  theme_solarized_2() +
  scale_colour_manual(values = c("#dc322f", "#268bd2"))
###### work in progress -- L/P budget last 5 years
```

### Auburn
```{r echo= FALSE, message = FALSE, warning = FALSE}
data_new <- data
data_new$City <- factor(data_new$City,
                        levels = c("Brunswick", "Lewiston", "Auburn", "Portland"))
###### create a vector to properly order towns

data_new %>%
  filter(City == 'Lewiston'| City == 'Portland'| City == 'Auburn') %>%
  ggplot(aes(`Fiscal Year`, `% on Police`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years) +
  scale_y_continuous("% of Budget spent on Police", limits = c(0.05, 0.20)) +
  stat_smooth(method = "lm", geom = "line", alpha = 0.75, size = 1, se = FALSE) +
  geom_point() +
  ggtitle("Police Spending in Lewiston/Portland/Auburn") +
  theme_solarized_2() +
  scale_colour_manual(values = c("#dc322f", "#859900", "#268bd2"))
###### work in progress - Lewiston/Auburn/Portland Police Budget
```

In Auburn, the difference is not as pronounced as the one with Portland.

Again, we reduce the scope to see a more recent picture

```{r echo= FALSE, message = FALSE, warning = FALSE}
data_new %>%
  filter(City == 'Lewiston'| City == 'Portland'| City == 'Auburn') %>%
  filter(`Fiscal Year` >= 2017) %>%
  ggplot(aes(`Fiscal Year`, `% on Police`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years, limits = c(2017, 2022)) +
  scale_y_continuous("% of Budget spent on Police", limits = c(0.05, 0.20)) +
  stat_smooth(method = "lm", geom = "line", alpha = 0.75, size = 1, se = FALSE) +
  geom_point() +
  ggtitle("Police Spending in Lewiston/Portland/Auburn, last 5 years") +
  theme_solarized_2() +
  scale_colour_manual(values = c("#dc322f", "#859900", "#268bd2"))
###### work in progress - L/A/P/ Budget last 5 years
```

### Brunswick

```{r echo= FALSE, message = FALSE, warning = FALSE}
data_new %>%
  filter(City == 'Brunswick'|City == 'Lewiston'| City == 'Portland'| City == 'Auburn') %>%
  ggplot(aes(`Fiscal Year`, `% on Police`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years) +
  scale_y_continuous("% of Budget spent on Police", limits = c(0.05, 0.20)) +
  stat_smooth(method = "lm", geom = "line", alpha = 0.75, size = 1, se = FALSE) +
  geom_point(alpha = 0.5) +
  ggtitle("Police Spending in Lewiston/Portland/Auburn/Brunswick") +
  theme_solarized_2() +
  scale_colour_manual(values = c("#6c71c4", "#dc322f", "#859900", "#268bd2"))
###### work in progress Lewiston/Portland/Auburn/Brunswick
```

In introducing Brunswick to the plot we see our highest value yet - 19.9% in FY 2010. However over the course of the last 12 years, the spending has been reduced down to 15.3% in FY 2022. 0.6% higher than Lewiston.

The reduced scope brings into focus the contrast between Lewiston's upward trajectory and the other town's downward trajectories.

```{r echo= FALSE, message = FALSE, warning = FALSE}
data_new %>%
  filter(City == 'Brunswick'|City == 'Lewiston'| City == 'Portland'| City == 'Auburn') %>%
  filter(`Fiscal Year` >= 2017) %>%
  ggplot(aes(`Fiscal Year`, `% on Police`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years, limits = c(2017, 2022)) +
  scale_y_continuous("% of Budget spent on Police", limits = c(0.05, 0.20)) +
  stat_smooth(method = "lm", geom = "line", alpha = 0.75, size = 1, se = FALSE) +
  geom_point(alpha = 0.5) +
  ggtitle("Police Spending in Lewiston/Portland/Auburn/Brunswick, last 5 years") +
  theme_solarized_2() +
  scale_colour_manual(values = c("#6c71c4", "#dc322f", "#859900", "#268bd2"))
###### work in progress Lewiston/Portland/Auburn/Brunswick last 5 years
```

## Spending Trajectories

```{r echo= FALSE, message = FALSE, warning = FALSE}
colors <- c ("#586e75", "#586e75", "#dc322f", "#586e75", 
             "#586e75", "#dc322f", "#586e75","#586e75","#dc322f")
###### color coded for increasing cities in the last 10 years (workaround)

allcities <- subset(data, select = c('City', 'Fiscal Year', '% on Police'))
###### dataset with just City, Fiscal Year, and % on Police

allcities %>%
  filter(City != 'Saco') %>%
  ggplot(aes(`Fiscal Year`, `% on Police`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years) +
  scale_y_continuous("% of Budget spent on Police", limits = c(0.05, 0.20)) +
  stat_smooth(method = "lm", geom = "line", alpha = 0.75, size = 1, se = FALSE) +
  geom_label(aes(x = 2016.5, y = 0.185, label = "Bangor")) +
  geom_label(aes(x = 2019, y = 0.1475, label = "Lewiston")) +
  geom_label(aes(x = 2014, y = 0.115, label = "Westbrook")) +
  ggtitle("Upward Police Spending Trajectories in Maine") +
  theme_solarized_2() +
  theme(legend.position = "none") +
  scale_colour_manual(values = colors)
###### work in progress Police Spending Trajectories red/black 
```

Of the 9 cities in the data set, 3 of them have upward spending trajectories by percentage - Westbrook, Bangor & Lewiston.

When the scope is reduced, 2 others join the fray - Augusta and South Portland

```{r echo= FALSE, message = FALSE, warning = FALSE}
colors2 <- c("#586e75","#dc322f", "#dc322f", "#586e75", "#586e75", "#dc322f",
             "#586e75", "#dc322f", "#dc322f" )
allcities %>%
  filter(City != 'Saco') %>%
  filter('Fiscal Year' >= 2017) %>%
  ggplot(aes(`Fiscal Year`, `% on Police`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years, limits = c(2017, 2022)) +
  scale_y_continuous("% of Budget spent on Police", limits = c(0.05, 0.20)) +
  stat_smooth(method = "lm", geom = "line", alpha = 0.75, size = 1, se = FALSE) +
  geom_label(aes(x = 2017.40, y = .181, label = "Bangor")) +
  geom_label(aes(x = 2018.4, y = .159, label = "Augusta")) +
  geom_label(aes(x = 2019.2, y = .144, label = "Lewiston")) +
  geom_label(aes(x = 2020, y = .13, label = "South Portland")) +
  geom_label(aes(x = 2021, y = 0.107, label = "Westbrook")) +
  ggtitle("Upward Police Spending Trajectories in Maine") +
  theme_solarized_2() +
  theme(legend.position = "none") +
  scale_colour_manual(values = colors2)
###### work in progress Police Spending Trajectories, last 5 years
```

In an effort to better illustrate the slope of these trajectory lines, I subtracted each FY 2021 budget percentage by the FY 2012 budget percentage for each town and plotted the values.

```{r echo= FALSE, message = FALSE, warning = FALSE}
compare_percentage <- na.omit(subset(data, select = c('City', '% Change 2012-2021'))) %>%
  filter(City != 'Saco')
###### create data set for comparing percentages for spending trajectories

compare_percentage %>%
  mutate(city = fct_reorder(City, compare_percentage$`% Change 2012-2021`)) %>%
  ggplot(aes(x = city, y = compare_percentage$`% Change 2012-2021`, label = City)) +
  geom_bar(stat = "identity", alpha = 0.6, width = 0.3, fill = "#dc322f") +
  scale_y_continuous("% Change", limits = c(-0.05, 0.05)) +
  ggtitle("Difference in Police Spending by Percentage 2012-2021") +
  scale_x_discrete("City") +
  theme_solarized_2() 
###### work in progress Difference in Police Spending by Percentage
```

Here it is with the FY 2017-2021 scope. Here we see Lewiston with the steepest increase in police spending as a percentage of the budget over this timespan.

```{r echo= FALSE, message = FALSE, warning = FALSE}
compare_percentage2 <- na.omit(subset(data, select = c('City', '% Change 2017-2021'))) %>%
  filter(City != 'Saco')
##### create dataset for comparing percentages from previous graph 2017-2021

compare_percentage2 %>%
  mutate(city = fct_reorder(City, compare_percentage2$'% Change 2017-2021')) %>%
  ggplot(aes(x = city, y = compare_percentage2$`% Change 2017-2021`, label = City)) +
  geom_bar(stat = "identity", alpha = 0.6, width = 0.3, fill = "#dc322f") +
  scale_y_continuous("% Change", limits = c(-0.05, 0.05)) +
  ggtitle("Difference in Police Spending by Percentage 2017-2021") +
  scale_x_discrete("City") +
  theme_solarized_2() 
###### work in progress Difference in Police Spending by Percentage last 5 years
```

## Police Spending Per Citizen

Another way to analyze police spending is to calculate the dollars spent per citizen. Here's a graph of Lewiston's police dollars spent per citizen.

``` {r echo= FALSE, message = FALSE, warning = FALSE}
Lewiston_data %>%
  ggplot(aes(`Fiscal Year`, `Police$ per Citizen`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years, limits = c(2010, 2020)) +
  scale_y_continuous("$ per Citizen") +
  stat_smooth(method = "lm", geom = "line", alpha = 0.75, size = 1, se = FALSE) +
  geom_point() +
  ggtitle("Police Spending Per Citizen") +
  theme_solarized_2() +
  scale_colour_manual(values = c("#dc322f"))
##### Lewiston $ per citizen graph
```

When adding in the 3 other towns from our earlier graph - Portland, Auburn & Brunswick - we see Lewiston at the bottom.

``` {r echo= FALSE, message = FALSE, warning = FALSE}
data_new %>%
  filter(City == 'Brunswick'|City == 'Lewiston'| City == 'Portland'| City == 'Auburn') %>%
  ggplot(aes(`Fiscal Year`, `Police$ per Citizen`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years, limits = c(2010, 2020)) +
  scale_y_continuous("$ per Citizen") +
  geom_smooth(method = "lm", alpha = 0.7, size = 1, se = FALSE) +
  geom_point(alpha = 0.5) +
  ggtitle("Police Spending Per Citizen in Lewiston/Portland/Auburn/Brunswick") +
  theme_solarized_2() +
  scale_colour_manual(values = c("#6c71c4", "#dc322f", "#859900", "#268bd2"))
##### Lewiston/Auburn/Brunswick/Portland Police $ per Citizen graph
```

In an effort to control for the differing income levels between towns, I plotted the dollar per citizen value divided by the median household income for each town and plotted those values. We can see Lewiston in the middle of this graph.

``` {r echo= FALSE, message = FALSE, warning = FALSE}
allcities2 <- na.omit(subset(data, select = c('City', 'Fiscal Year', 'Police $ per income')))
##### data set for all cites police/income comparison

colors3 <- c("#586e75","#586e75", "#586e75", "#586e75", "#586e75", "#dc322f",
             "#586e75", "#586e75", "#586e75" )
##### color set to highlight Lewiston

allcities2 %>%
  filter(City != 'Saco') %>%
  ggplot(aes(`Fiscal Year`, `Police $ per income`, color = City)) +
  scale_x_continuous("Fiscal Year", breaks = years) +
  scale_y_continuous("Police $ per Citizen, divided by Median Income") +
  geom_smooth(method = "lm", size = 1, se = FALSE) +
  ggtitle("Police Spending Per Citizen Divided by Median Income") +
  geom_label(aes(x = 2010.3, y = .0043, label = "Lewiston")) +
  geom_label(aes(x = 2012.1, y = 0.0062, label = "Augusta")) +
  geom_label(aes(x =  2014.2, y = 0.00735, label = "Bangor")) +
  geom_label(aes(x = 2015.8, y = 0.0026, label = "South Portland")) +
  geom_label(aes(x = 2017.8, y = 0.00335, label = "Westbrook")) +
  geom_label(aes(x = 2014.3, y = 0.00548, label = "Brunswick")) +
  theme_solarized_2() +
  theme(legend.position = "none") +
  scale_colour_manual(values = colors3)
##### comparison of all cities while accounting for median income
```